\thetitle

### 2009-02-10 paper/paper.tex
paper: Fix awkward code layout in M_ne_tree definition.
Ignore-this: 305cffecce046ade027e39effdb3533f

darcs-hash:20090210133833-bab43-dd4d4e0c5fa9d52d579fbcc8faa16639afad69f6.gz
paper: Make some formulas slightly less cramped by putting recursive qs calls on separate lines.
Ignore-this: ede624d932bb56b09857f59561f4a7aa

darcs-hash:20090210142447-bab43-5a518aad267ccce815c17512d101547dabefe1b5.gz
paper: Add paragraph about existing work; address some other referee comments.
Ignore-this: fd7b93d1630b9b83a22495ffdcf67daa

darcs-hash:20090210142617-bab43-447e2bf27f361e2576ceac342a8023fb390b2dd2.gz

A completely different approach to type-theoretic analysis of computational complexity is to devise a special-purpose type theory in which the types of terms include some form of complexity guarantees. This is the approach taken in \cite{constable}, for example. However, as far as we can see, this method is not suitable for the formalization of a traditional complexity proof in which an algorithm is first defined and then separately analyzed. We are not aware of practical techniques usable for this purpose apart from the one we present in this paper.

@@

Combined, these two traits ensure that the $O(n \log n)$ average-case bound holds not just averaged over all possible input lists, but for each individual input list as well. This greatly simplifies things, because it means that if we prove the bound for an arbitrary input, the global bound immediately follows.

Hence, we now modify our |qs| definition to use a three-way partition pass, and to use nondeterministic pivot selection. Following that, we will compose a new monad (again using |MMT|, but this time transforming a nondeterminism monad) that the new definition can be instantiated with to capture its \emph{expected} comparison count.

@@

Note that the Quicksort definition in section \ref{embed} could have been implemented with a |partition| pass as well, which might well have made the worst-case proof even simpler. We chose not to do this, in order to emphasize that the properties the average-case proof demands of the algorithm rule out the naive but familiar implementation using |filter| passes.

@@

  Fixpoint bind_ne_tree (A B: Set) (m: ne_tree A) (k: A -> ne_tree B): ne_tree B :=
    match m with
    | Leaf a => k a
    | Node ts => Node (ne_list.map (fun x => bind_ne_tree x k) ts)
    end.
\end{code}
\begin{code}
  Definition M_ne_tree: Monad := Build_Monad ne_tree bind_ne_tree ret_ne_tree.

@@

  Definition qs_U: list nat -> list nat := qs U cmp_U pick_U.
\end{code}
|qs_U| operates directly on lists of indices into |sort l|. Comparison of indices is defined by comparison of the |T| values they denote in |sort l|. Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of indices compared, by using |MMT| with |Monoid_U|, the free monoid over |nat * nat| pairs, instead of the additive monoid on |nat| we used up until now.

@@

\begin{code}
  forall (X: Set) (fr: X -> R) (q: list X) (t: ne_tree (list X)),
    (forall x `In` q, expec (count x) t <= fr x) ->
    (forall x `NotIn` q, expec (count x) t = 0) -> expec length t <= {-"\displaystyle{\sum_{\Varid{x} \in \Varid{q}}\Varid{fr}\ \Varid{x}}"-}.
\end{code}
We end up with two subgoals, the first of which is \begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using  the last of the identities listed for |expec| in section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in
\begin{code}
  forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0,
\end{code}
which can be shown by induction on |qs|, although we will not do so in this paper. We \emph{will} use this property again in the next section.

>>>>

\paragraph{Related work}

In his Ph.D thesis \cite{hurd}, Hurd presents an approach to formal analysis of probabilistic programs based on a comprehensive formalization of measure-theoretic constructions of probability spaces, representing probabilistic programs using a state-transforming monad in which bits from an infinite supply of random bits may be consumed. In \cite{mohringaudebaud}, Audebaud and Paulin-Mohring describe a different monadic approach in which programs are interpreted directly as measures representing probability distributions. A set of axiomatic rules is defined for estimating the probability that programs interpreted this way satisfy certain properties.

Compared to these approaches, our infrastructure for reasoning about nondeterministic programs is far less ambitious, supporting only a very simplistic notion of expectation, computed simply by monadically collecting all possible outputs (which we only permit to be finitely many), and then taking the average of a measure applied to these outputs. In particular, we do not concern ourselves at all with probability distributions.

A completely different approach to type-theoretic analysis of computational complexity is to devise a special-purpose type theory in which the types of terms include some form of complexity guarantees. This is the approach taken in \cite{constable}, for example.

@@

Combined, these two traits ensure that the $O(n \log n)$ average-case bound holds not just averaged over all possible input lists, but for each individual input list as well. This greatly simplifies things, because it removes the need to carefully track input distributions in order to show that nice inputs (for which the original deterministic version of the algorithm performs well) sufficiently outnumber bad inputs (for which the original deterministic version of the algorithm performs poorly). In particular, it means that once we prove that the bound holds for an arbitrary input, the global bound immediately follows. It also means that for a key lemma near the end of our proof, we can use straightforward induction over the algorithm's recursive call structure, without having to show that given appropriately distributed inputs, the partition step yields lists that are again appropriately distributed. The latter is a typical concern in more ambitious approaches for average-case complexity analysis \cite[for example]{modularcalculus} and analysis of probabilistic algorithms.

While we sympathize with the view that changing the algorithm under investigation amounts to cheating, we will nevertheless simply follow the textbook proof (which represents common practice) in this regard. Hence, we now modify our |qs| definition to use a three-way partition pass, and to use nondeterministic pivot selection. Following that, we will compose a new monad (again using |MMT|, but this time transforming a nondeterminism monad) that the new definition can be instantiated with to capture its \emph{expected} comparison count.

@@

Note that the deterministic Quicksort definition in section \ref{embed} could have been implemented with a |partition| pass as well, which might well have made the worst-case proof even simpler. We chose not to do this, in order to emphasize that the properties the average-case proof demands of the algorithm rule out the naive but familiar implementation using |filter| passes.

@@

  Fixpoint bind_ne_tree (A B: Set)
    (m: ne_tree A) (k: A -> ne_tree B): ne_tree B :=
      match m with
      | Leaf a => k a
      | Node ts => Node (ne_list.map (fun x => bind_ne_tree x k) ts)
      end.

  Definition M_ne_tree: Monad := Build_Monad ne_tree bind_ne_tree ret_ne_tree.

@@

  Definition qs_U: list nat -> list nat := qs U cmp_U pick_U.
\end{code}
|qs_U| operates directly on lists of indices into |sort l|. Comparison of indices is defined by comparison of the |T| values they denote in |sort l|. Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of indices compared, by using |MMT| with |Monoid_U|, the free monoid over |nat prod nat| pairs, instead of the additive monoid on |nat| we used up until now.

@@

\begin{code}
  forall (X: Set) (fr: X -> R) (q: list X) (t: ne_tree (list X)),
    (forall x `In` q, expec (count x) t <= fr x) ->
    (forall x `NotIn` q, expec (count x) t = 0) -> expec length t <= {-"\displaystyle{\sum_{\Varid{x} \in \Varid{q}}\Varid{fr}\ \Varid{x}}"-}.
\end{code}
We end up with two subgoals, the first of which is \begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using identity \ref{expec_bind_lem} from section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in
\begin{code}
  forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0,
\end{code}
which can be shown by induction on |qs|, although we will not do so in this paper. We \emph{will} use this property again in the next section.
